A Horrocks-type theorem for even orthogonal $K_2$

### 2019-08-21 dl-paper.tex
Added more citations

\documentclass[oneside, 8pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, mathtools, xfrac, microtype, nameref, thmtools}
\usepackage[breaklinks=true,unicode]{hyperref}
\usepackage[hyperref=true, backend=biber, bibencoding=utf8, giveninits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage[notref,notcite]{showkeys}

@@

\begin{document}
\begin{abstract} We prove an analogue of Horrocks theorem for even-dimensional orthogonal Steinberg groups. This result transfers to the orthogonal case an earlier result of M.~Tulenbaev and is also a $\K_2$-analogue of an earlier result of A.~Suslin and V.~Kopeiko. \end{abstract}
\maketitle
\section{Introduction}
The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the brilliant achievements of mathematics of the 70's, which bolstered the development of modern algebraic K-theory. In its weakest form Serre problem asks whether any projective module over a polynomial ring $R = k[x_1,\ldots, x_n]$ over a field $k$ is free. We refer the reader to the textbook~\cite{Lam10} for a comprehensive account on the problem, its history and the subsequent affirmative solution.

After the original Serre problem had been solved, numerous analogous statements have drawn the attention of specialists. For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_{1,n}(R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has trivial values on polynomial rings $k[x_1, \ldots x_n]$ over fields for all $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently, A.~Stavrova has obtained probably the most general result in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on arbitary isotropic reductive group scheme $G/k$ of isotropic rank $\geq 2$ (see~\cite[Theorem~1.2]{St-poly}).

@@

One might ask whether any analogue of Serre's problem also exists for the functor $\K_2$ and it turns out that it is indeed so. More precisely, in~\cite{Tu83} M. Tulenbaev demonstrates an ``early stability theorem'' from which the isomorphism $\K_2(\rA_\ell, k[x_1, \ldots x_n]) \cong \K_2(\rA_\ell, k) = \K^M_2(k)$ immediately follows for $\ell \geq 4$. While a number of new results on the $\K_1$-analogue of Serre's problem have appeared in the literature since~\cite{Su77} (see e.\,g.~\cite{Abe83, St-poly}), little progress has been made for the $\K_2$-analogue since~\cite{Tu83}. Currently the conjecture still remains open for all $\Phi$ different from $\rA_\ell$, $\ell \geq 4$.

@@

\subsection{Acknowledgements} Theorem 2 is obtained by the first-named author who acknowledges the financial support of RSCF grant No ???.
Theorem 3 is obtained by the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044

@@

\section{An injectivity theorem for Steinberg groups} \label{firstPart}
We start this section by recalling basic notation and facts pertaining to the theory of Grothendieck--Witt groups.
%First of all, recall that the stable orthogonal K-groups $\KO_i(R) = \KO_i(\infty, R)$ are precisely the hermitian K-groups ${}_{-1}\!L_i(R)$ introduced by Karoubi in~\cite{Ka80}. We will need yet another redesignation for these groups $\KO_i(R)$, which comes from the theory of {\it higher Grothendieck--Witt groups of schemes}.
Recall that this theory, developed by M.~Schlichting, is a modern broad generalization of the classical hermitian K-theory of rings (see e.\,g.~\cite{Ka80}). We refer the reader to~\cite[\S~2]{FRS12} and~\cite[\S~2]{AF17} for an introduction to Grothendieck--Witt groups.

@@

\begin{theorem} \label{thm41}
Suppose that $2 \in A^*$. Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective.
\end{theorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following computation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

\begin{rem} For a simply-laced $\Phi$~\cref{thm:Stepanov} can be deduced from~\cref{thm:Tits} by means of~\cref{Zrels}. Indeed, consider the operator $d \colon 2^\Phi \to 2^\Phi$ of root subsets given by $d(U) = U \cup (U - U)\cap \Phi$, $U \subseteq \Phi$. In other words, $d$ adjoins to $U$ all differences of roots from $U$ which are themselves roots. It is not hard to show that for any parabolic subset $S \subseteq \Phi$ the subset $\Sigma_S$ has the property that $d^n(\Sigma_S) = \Phi$ (in fact, $n=2$). It remains to see that relation~\eqref{Z5} immediately implies that every group $G$ containing $\mathcal{Z}(U, R, I)$ also contains $\mathcal{Z}(dU, R, I)$. \end{rem}

@@

We are ready for the first main result of this section.
\begin{theorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{theorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g \in G_M^0$ and consequently $g = 1$.

>>>>

\documentclass[oneside, 8pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, mathtools, xfrac, microtype, nameref, thmtools}
\usepackage[breaklinks=true,unicode]{hyperref}
\usepackage[hyperref=true, backend=biber, bibencoding=utf8, giveninits=true, citestyle=numeric-comp, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}
\usepackage[capitalize]{cleveref}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{tikz}
\usepackage{enumitem}
%\usepackage[notref,notcite]{showkeys}

@@

\begin{document}
\begin{abstract} We prove an analogue of Horrocks theorem for even-dimensional orthogonal Steinberg groups. This result transfers to the orthogonal case an earlier result of M.~Tulenbaev and is also a $\K_2$-analogue of an earlier result of A.~Suslin and V.~Kopeiko. \end{abstract}
\maketitle
\section{Introduction}
The solution of the famous Serre problem on projective modules by D.~Quillen and A.~Suslin was one of the brilliant achievements of mathematics of the 70's, which bolstered the development of modern algebraic K-theory, see~\cite{Su76, Qu76}. In its weakest form Serre problem asks whether any projective module over a polynomial ring $R = k[x_1,\ldots, x_n]$ over a field $k$ is free. We refer the reader to the textbook~\cite{Lam10} for a comprehensive account on the problem, its history and the subsequent affirmative solution.

After the original Serre problem had been solved, numerous analogous statements have drawn the attention of specialists. For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_{1,n}(R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has trivial values on polynomial rings $k[x_1, \ldots x_n]$ over fields for all $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently, A.~Stavrova has obtained probably the most general results in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on arbitary isotropic reductive group scheme $G$ of isotropic rank $\geq 2$ (see~\cite[Theorem~1.2]{St-poly} for the result over a field and ~\cite[Corollary~1.2]{St-Ded} for its further generalization to Dedekind domains of arithmetic type).

@@

One might ask whether any analogue of Serre's problem also exists for the functor $\K_2$ and it turns out that it is indeed so. More precisely, in~\cite{Tu83} M. Tulenbaev demonstrates an ``early stability theorem'' from which the isomorphism $\K_2(\rA_\ell, k[x_1, \ldots x_n]) \cong \K_2(\rA_\ell, k) = \K^M_2(k)$ immediately follows for $\ell \geq 4$. While a number of new results on the $\K_1$-analogue of Serre's problem have appeared in the literature since~\cite{Su77} (see e.\,g.~\cite{Abe83, St-poly, St-Ded}), little progress has been made for the $\K_2$-analogue since~\cite{Tu83}. Currently the conjecture still remains open for all $\Phi$ different from $\rA_\ell$, $\ell \geq 4$.

@@

\subsection{Acknowledgements} \cref{thm41} is obtained by the first-named author who acknowledges the financial support of RSCF grant No ???.
\cref{thm:P1glueing} is obtained by the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044

@@

\section{An injectivity theorem for Steinberg groups} \label{firstPart}
We start this section by recalling basic notation and facts pertaining to the theory of Grothendieck--Witt groups.
%First of all, recall that the stable orthogonal K-groups $\KO_i(R) = \KO_i(\infty, R)$ are precisely the hermitian K-groups ${}_{-1}\!L_i(R)$ introduced by Karoubi in~\cite{Ka80}. We will need yet another redesignation for these groups $\KO_i(R)$, which comes from the theory of {\it higher Grothendieck--Witt groups of schemes}.
Recall that this theory, developed by M.~Schlichting, is a modern broad generalization of the classical hermitian K-theory of rings (see~\cite{Ka80}). We refer the reader to~\cite[\S~2]{FRS12} and~\cite[\S~2]{AF17} for an introduction to Grothendieck--Witt groups.

@@

\begin{externaltheorem} \label{thm41} Suppose that $2 \in A^*$. Then for $\ell \geq 7$ the canonical map $\St(\rD_\ell, B) \to \St(\rD_\ell, R)$ is injective. \end{externaltheorem}
\begin{proof}
 Consider the following commutative diagram with exact rows, in which the lifting $t$ is obtained from~\cref{lem:lemma32}:
\begin{equation*} \xymatrix{
 C(\rD_\ell, B, I) \ar[r]^{\lambda_B} \ar@{->>}[d]_{f} & \St(\rD_\ell, B, I) \ar[r]^{\mu_B} \ar[d]_{g} &
 \St(\rD_\ell, B) \ar[r]^{\nu_B} \ar[d]_{h} & \St(\rD_\ell, k[X]) \ar[d]_{i} \\
 C(\rD_\ell, R, I) \ar[r]^{\lambda_R}         & \St(\rD_\ell, R, I) \ar[r]^{\mu_R} \ar@{-->}[ur]^{t}&
 \St(\rD_\ell, R) \ar[r]^-{\nu_R}        & \St(\rD_\ell, k[X, X^{-1}]).
}\end{equation*}
Let $a$ be an element of $\Ker(h)$. Since $i$ is injective by~\cref{field-injectivity}, the element $a$
 also lies in $\Ker(\nu_B)$ and hence comes from some $b \in \St(\rD_\ell, B, I)$ via $\mu_B$.
Since $g(b) \in \Ker(\mu_R)$ there exists some $c \in C(\rD_\ell, R, I)$ such that $\lambda_R(c) = g(b)$. 
By~\cref{lem:prop41}, $f$ is surjective, therefore $c = f(d)$ for some $d \in C(\rD_\ell, R, I)$.
The required assertion now follows from the following computation:
 \[ 1 = \mu_B\lambda_B(d) = tg\lambda_B(d) = t\lambda_Rf(d) =t(g(b)) = \mu_B(b) = a. \qedhere \]
\end{proof}

@@

\begin{rem} For a simply-laced $\Phi$~\cref{thm:Stepanov} can be deduced from~\cref{thm:Tits} by means of~\cref{Zrels}. Indeed, consider the operator $d \colon 2^\Phi \to 2^\Phi$ of root subsets given by $d(U) = U \cup (U - U)\cap \Phi$, $U \subseteq \Phi$. In other words, $d$ adjoins to $U$ all differences of roots from $U$ which are themselves roots. It is not hard to show that for any parabolic subset $S \subseteq \Phi$ the subset $\Sigma_S$ has the property that $d^n(\Sigma_S) = \Phi$ for some $n>1$ (in fact, $n=2$). It remains to see that relation~\eqref{Z5} immediately implies that every group $G$ containing $\mathcal{Z}(U, R, I)$ also contains $\mathcal{Z}(dU, R, I)$. \end{rem}

@@

We are ready for the first main result of this section.
\begin{externaltheorem} \label{thm:P1glueing}
 Let $\Phi$ be a root system of type $\rA_{\geq 4}, \rD_{\geq 5}$ or $\rE_{6,7,8}$.
 Suppose that the map $j_R \colon \St(\Phi, B) \to \St(\Phi, A[X, X\inv])$ is injective.  
 Then the map $j_-$ is injective and the following commutative square is pullback
 \begin{equation} \nonumber
  \xymatrix{ \St(\Phi, A) \ar[r] \ar[d] & \St(\Phi, A[X]) \ar[d]^{j_+} \\ \St(\Phi, A[X\inv]) \ar[r]^{j_-} & \St(\Phi, A[X, X\inv]).}
 \end{equation}
\end{externaltheorem}
\begin{proof} Since $j_- = j_R j_B^-$, where $j_B^-$ is as in~\cref{prop43}, it suffices to show that $j_B^-$ is injective.
 Set $v_0 = [1, 1, 1] \in \overline{V}_T$ and suppose $g \in \Ker(j_B)$.
 By~\cref{prop43} holds $v_0 = j_B^-(g) \cdot v_0 = [1, g, 1]$ therefore $g \in G_M^0$ and consequently $g = 1$.
